---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Jacob Freedman"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
setwd("~/R/EDA-Fall2022")
getwd()
library(tidyverse)
library(lubridate)
library(agricolae)
library(formatR)


LakeChemPhys_Raw <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

LakeChemPhys_Raw$sampledate <- as.Date(LakeChemPhys_Raw$sampledate, format = "%m/%d/%y")
class(LakeChemPhys_Raw$sampledate)

#2
theme1 <- theme_classic(base_size = 12) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "left", axis.line = element_line(arrow = arrow()))

theme_set(theme1)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The July lake temperature, regardless of lake, does not vary with depth. 
Ha: The July lake temperature, regardless of lake, is related to depth. 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

LakeChemPhys_Filtered <- 
  mutate(LakeChemPhys_Raw, month = month(sampledate)) %>%
  filter(month == 7) %>%
  select(c(`lakename`, `year4`, `daynum`, `depth`, `temperature_C`)) %>%
  drop_na()
  

#5
Temp.Depth <- ggplot(LakeChemPhys_Filtered, aes(x=depth, y=temperature_C)) +
  geom_point() +
  geom_smooth(method="lm") +
  xlab("Depth (m)") +
  ylab("Temperature (C)") +
  ylim(0,35)
  print(Temp.Depth)
  
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: Based on the figure alone (not the regression analysis), there appears to be a relationship between temperature and depth, but it may not be linear. Variability in temperature is high between 0 and 6 m (but decreases overall), and then temperature slowly approaches 0 degrees. This suggests there may be some sort of exponential relationship, as temperature asymptotes towards 0 degrees as depth increases. 

7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
temp.regression <- lm(data = LakeChemPhys_Filtered, temperature_C ~ depth)
summary(temp.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: Based on the results of the model, there is a strong relationship between temperature and depth. Depth explains 73.87% of variability in temperature, based on 9726 degrees of freedom. The model shows this relationship is statistically significant, with a p value < 2.2e-16. Temperature is expected to decrease by 1.95 degrees for every 1m increase in depth.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
AIC <- lm(data = LakeChemPhys_Filtered, temperature_C ~ year4 + daynum + depth)
summary(AIC)

#All 3 variables appear t0 have strong relationships with temperature, so we will consider them all and run a stepwise function to confirm.

step(AIC) 

#The stepwise function does not recommend removing any of the variables from the model.

#10
Tempmodel <- lm(data = LakeChemPhys_Filtered, temperature_C ~ year4 + daynum + depth)
summary(Tempmodel)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggests we use year, daynum, and depth to predict temperature. These variables predict 74.11% of variability in temperature, which is a very small improvement compared to depth alone (73.87%).

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
Laketemp.anova <- aov(data = LakeChemPhys_Filtered, temperature_C ~ lakename)
summary(Laketemp.anova)

Lakedepth.lm <- lm(data = LakeChemPhys_Filtered, temperature_C ~ lakename)
summary(Lakedepth.lm)


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference in mean temperature among the lakes. Both the lm and anova showed p values of <2.2e-16, which means that there are at least two lakes with differences in mean temperature. We will need post hoc tests to determine which lakes have different mean temperatures from one another. 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r, fig.height = 6, fig.width = 8}
#14.
LakeTemps <- ggplot(LakeChemPhys_Filtered, aes(x=depth, y=temperature_C, color = lakename)) +
  geom_point(alpha = 0.5, cex = 0.4) +
  geom_smooth(method="lm", se = FALSE) +
  xlab("Depth (m)") +
  ylab("Temperature (C)") +
  ylim(0,35) +
  theme(legend.position = "left", legend.text = element_text(size = 10))
  print(LakeTemps)


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Laketemp.anova)

Laketemp.groups <- HSD.test(Laketemp.anova, "lakename", group = TRUE)
Laketemp.groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Paul and Ward lake did not have statistically different mean temperatures from Peter Lake (p=0.22, 0.78 respectively). According to our groups analysis, no lakes had mean temperatures statistically distinct from all other lakes. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could use a T test because we are comparing the means of just two populations (derived from two samples). 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
LakesCramptonWard <- 
  mutate(LakeChemPhys_Raw, month = month(sampledate)) %>%
  filter(month == 7, (lakename == "Crampton Lake" | lakename == "Ward Lake")) %>%
  select(c(`lakename`, 'sampledate', `year4`, `daynum`, `depth`, `temperature_C`)) %>%
  drop_na()

Temperature.twosample <- t.test(LakesCramptonWard$temperature_C ~ LakesCramptonWard$lakename)
Temperature.twosample
```

>Answer: The T test showed there is not a significant difference in the mean temperature of Crampton and Ward Lake (p=0.26). This is in agreement with our prior groups analysis, where Ward and Crampton Lake were both found to be in group b. 
