---
title: "Assignment 09: Data Scraping"
author: "Jacob Freedman"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A09_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1
setwd("~/R/EDA-Fall2022")
getwd()
library(tidyverse)
library(rvest)
library(lubridate)

theme1 <- theme_classic(base_size = 12) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "left", axis.line = element_line(arrow = arrow()))

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2021 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2
DurhamWaterWebpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021')
DurhamWaterWebpage

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "27.6400".

```{r scrape.the.data}
#3
water_system_name <- DurhamWaterWebpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
water_system_name

pwsid <- DurhamWaterWebpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
pwsid
  
ownership <- DurhamWaterWebpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
ownership
  
max_withdrawals_mgd <- DurhamWaterWebpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
max_withdrawals_mgd

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc...

5. Create a line plot of the maximum daily withdrawals across the months for 2021

```{r create.a.dataframe.from.scraped.data}
#4
df_max_withdrawals <- data.frame("Month_abbr" = c("Jan", "May", "Sep", "Feb", "Jun", "Oct", "Mar", "July", "Nov", "Apr", "Aug", "Dec"),
                             "Month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
                             "Year" = rep(2021,12),
                             "Max_withdrawals_mgd" = as.numeric(max_withdrawals_mgd))

df_max_withdrawals <- df_max_withdrawals %>% 
  mutate(Ownership = !!ownership,
         PWSID = !!pwsid,
         Water_System = !!water_system_name,
         Date = my(paste(Month,"-",Year))) %>%
         arrange(Month)

#5
Max_withdrawl_plot <- ggplot(df_max_withdrawals,aes(x=Date, y=Max_withdrawals_mgd)) +
  geom_line() +
  theme1 +
  ylab("Maximum Monthly Withdrawl")
print(Max_withdrawl_plot)


```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6.
the_base_url <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php'
pwsid <- '03-32-010'
the_year <- 2019
Scrape_url <- paste0(the_base_url, '?pwsid=', pwsid, '&year=', the_year)
print(Scrape_url)

#Retrieve the website contents 
the_website <- read_html(Scrape_url)

Scrape_function <- function(the_year, pwsid){
  
  The_website <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php', '?pwsid=', pwsid, '&year=', the_year))

  water_system_tag <- "div+ table tr:nth-child(1) td:nth-child(2)"
  PWSID_tag <- "td tr:nth-child(1) td:nth-child(5)"
  Ownership_tag <- "div+ table tr:nth-child(2) td:nth-child(4)"
  Max_withdrawal_tag <- "th~ td+ td"

  water_system_name <- The_website %>% html_nodes(water_system_tag) %>% html_text()
  pwsid <- The_website %>% html_nodes(PWSID_tag) %>%  html_text()
  ownership <- The_website %>% html_nodes(Ownership_tag) %>% html_text()
  max_withdrawals_mgd <- The_website %>% html_nodes(Max_withdrawal_tag) %>% html_text()

  df_max_withdrawals <- data.frame("Month_abbr" = c("Jan", "May", "Sep", "Feb", "Jun", "Oct", "Mar", "July", "Nov", "Apr", "Aug", "Dec"),
                             "Month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
                             "Year" = rep(the_year,12),
                             "Max_withdrawals_mgd" = as.numeric(max_withdrawals_mgd))

  df_max_withdrawals <- df_max_withdrawals %>% 
  mutate(Ownership = !!ownership,
         PWSID = !!pwsid,
         Water_System = !!water_system_name,
         Date = my(paste(Month,"-",Year))) %>%
         arrange(Month)

  
  Sys.sleep(1)

 return(df_max_withdrawals)
}


```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
Scraped_df <- Scrape_function(2015,'03-32-010')
view(Scraped_df)

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8
Scraped_df_Asheville <- Scrape_function(2015,'01-11-010')
Scraped_df_Durham <- Scrape_function(2015,'03-32-010')

Asheville_Durham_combined <- rbind(Scraped_df_Asheville, Scraped_df_Durham)

Withdrawl_plot_comparison <- ggplot(Asheville_Durham_combined,aes(x=Date, y=Max_withdrawals_mgd, color=Water_System)) +
  geom_line() +
  theme1 +
  ylab("Maximum Monthly Withdrawal")
print(Withdrawl_plot_comparison)
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

>TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data}
#9

Years2010to2019 <- c(2010:2019)
pwsidlist <- rep("01-011-010",10)

#"Map" the "scrape.it" function to retrieve data for all these
dfs_Asheville_2010to2019 <- map2(Years2010to2019, pwsidlist, Scrape_function) 

#Conflate the returned list of dataframes into a single one
dfs_combined_Asheville_2010to2019 <- rbind(dfs_Asheville_2010to2019)

Scraped_df_Asheville_2010 <- Scrape_function(2010,'01-11-010')
Scraped_df_Asheville_2011 <- Scrape_function(2011,'01-11-010')
Scraped_df_Asheville_2012 <- Scrape_function(2012,'01-11-010')
Scraped_df_Asheville_2013 <- Scrape_function(2013,'01-11-010')
Scraped_df_Asheville_2014 <- Scrape_function(2014,'01-11-010')
Scraped_df_Asheville_2015 <- Scrape_function(2015,'01-11-010')
Scraped_df_Asheville_2016 <- Scrape_function(2016,'01-11-010')
Scraped_df_Asheville_2017 <- Scrape_function(2017,'01-11-010')
Scraped_df_Asheville_2018 <- Scrape_function(2018,'01-11-010')
Scraped_df_Asheville_2019 <- Scrape_function(2019,'01-11-010')

Scraped_df_Asheville_2010to2019 <- rbind(Scraped_df_Asheville_2010, Scraped_df_Asheville_2011, Scraped_df_Asheville_2012, Scraped_df_Asheville_2013, Scraped_df_Asheville_2014, Scraped_df_Asheville_2015, Scraped_df_Asheville_2016, Scraped_df_Asheville_2017, Scraped_df_Asheville_2018, Scraped_df_Asheville_2019)

view(Scraped_df_Asheville_2010to2019)

Asheville_2010to2019_plot <- ggplot(Scraped_df_Asheville_2010to2019,aes(x=Date, y=Max_withdrawals_mgd)) +
  geom_smooth() +
  geom_line() +
  theme1 +
  ylab("Maximum Monthly Withdrawal")
print(Asheville_2010to2019_plot)

```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?

The maximum monthly water usage in Asheville appears stable from 2010 - 2017, but jumps dramatically in 2018 and 2019. Perhaps 2018 and 2019 were particularly dry years for the region and the city required more water use to meet constituents' needs. 